Standardization of ceramic shape: A case studay from the Iron Age pottery from northeastern Taiwan

### 2020-07-31 analysis/paper/paper.Rmd
edit the results and discussion of shape analysis

Key questions in archaeological shape analysis normally involve measuring shape standardization over time, or between geographical areas. Standardization is often investigated using multivariate analysis of shape variables computed from landmark data, or coefficients of variation on associated metric data, especially for lithic assemblages. For example, @Archer2015's case study of stone points in Southern Africa suggests an increase in shape standardization over time that may relate to increased maintenance of finished points. @Buchanan2018 studied metric data and GMM to identify a more uniform base-shape of Folsom points than Clovis points across the western US [@Buchanan2019]. Also using metric data and GMM, @Smith2016 found standardized bases of fluted point in Alaska and northern Yukon that might indicate a risk management strategy to ensure the ease of replacement during long-distance travel. Other factors, such as low levels of cultural innovation in a small group, could also lead to an increase in standardization of point shapes [@Okumura2014].

@@

As a standardized measure of the spread of data, coefficients of variation (CV) allows us to directly compare variation in samples measured with different units or means. This is useful to compare the degree of standardization for archaeological assembles and enables comparison of variation across different sample sizes [@Eerkens2001, pp.498]. Following @Eerkens2001 and @Roux2003, we take this as our measurement of standardization in vessel shape variables: lower CV values reflect higher standardization, and thus increased craft specialization in the community. Given that CV is robust for positive values due to the representation by ratio, we normalized PC scores to a range between 1 and 2 before computation of CV.

@@

We also conducted size analysis by examining the body diameter of the vessels to explore any differences in vessel form reflected by size and shape. We measured body diameter directly from each physical vessel in the collection, and we focus on this metric because it is available for more vessels than any other metric. The body diameter of vessels from the Chinese period is larger than those from the two earlier periods, and the vessels before European contact have the smallest body diameter on average (Figure \@ref(fig:size-shape-regression-plots): A). Our linear regression models (Figure \@ref(fig:size-shape-regression-plots): B) show that shape, captured by PC1 and PC2, changes with size across the three phases, except for PC1 in the Chinese period. That is, the shorter vessels are larger in body diameter according to the significant positive correlation between PC1 and size. However, vessels from the Chinese period does not show this pattern. On the contrary, the negative relationship between PC2 and size suggests that vessels with a narrower neck and mouth tend to have a larger body diameter.

@@

```{r size-analysis-prep-2}
# combine with cv information 
size_pcs_cv_join <-
  jpgs_imported_coo_size_pcs_metric %>% 
  group_by(y) %>% 
  mutate(Diameter1_Body_ave = round(mean(Diameter1_Body), 2),
         cv_body_diameter = round(sd(Diameter1_Body)/mean(Diameter1_Body), 
                                  digits = 2)*100) %>%
  mutate(phase = as.factor(y)) %>% 
  left_join(cv_pcs_all) %>% 
  left_join(pc1_df_with_filename)
  #distinct(phase, cv_pc1, cv_pc2, cv_pc3, Diameter1_Body_ave, cv_body_diameter)
```

```{r boxplot-size-by-phase}
size_cv_df <-
  size_pcs_cv_join %>% 
  distinct(phase, Diameter1_Body_ave, cv_body_diameter, cv_pc1) %>% 
  mutate(cv_body_diameter_label = paste("CV =", cv_body_diameter, "%"),
         Diameter1_Body_ave_label = paste("mean=", Diameter1_Body_ave))

@@

```{r boxplot-pc1-ave-size}
library(ggrepel)
y_limits <- c(NA, 32)
boxplot_ave_size_cv_pc <-
  size_cv_df %>% 
  ggplot(aes(cv_body_diameter, cv_pc1)) +
  geom_point() +
  geom_text_repel(label = size_cv_df$phase, 
                  size = 2.5) + 
  labs(x = "CV of body diameter (%)",
       y = "CV of PC1\n(%)") +
  theme(legend.position = "none") +
  theme_bw() +
  theme(axis.text.x = element_text(size = rel(0.9)),
        axis.text.y = element_text(size = rel(0.9)))
```

label = size_cv_df$phase,
                   size = 2.5, ylim = y_limits

@@

```{r size-regression-plot-together}
# plot boxplot and regression together 
left_column <-
  plot_grid(boxplot_size, boxplot_ave_size_cv_pc, 
            ncol = 1, rel_heights = c(1.1,1), 
            labels = c("A", "B"), label_size = 12 )

@@

ggsave(plot = size_boxplot_regression,
       here("analysis", "figures",
            "size-boxplot-regression.jpg"),
       width = 180, 
       height = 100,
       dpi = 300,
       units = "mm")
```

```{r, size-shape-regression-plots, echo=FALSE, fig.cap= "A: Distribution of the body diameter of the vessels across three phases. B: Scatter plots of shape variable, PC1 at the top and PC2 at the bottom, as a function of body diamater"}
knitr::include_graphics(here("analysis/figures/size-boxplot-regression.jpg"))
```

@@

Previous investigations at Kiwulan suggested an unequal distribution of prestige goods, trade ornaments in particular, following the appearance of Europeans [@Cheng2008; @Wang2011], which hinted at the emergence of social inequality within the Indigenous community. Ceramic vessel shape standardization was examined to measure craft specialization as a proxy for social differentiation [@Costin2001; @Junker1999]. The result of our MANOVA demonstrates a significant difference in shapes between the pre-European and Post-European periods, and between the pre-European and the Chinese periods. The average shape presents as a round body with a wide rim and neck before European contact, which shifts to a more oval-shaped body with narrower rim and neck after the European presence. Shape changes are more pronounced during the Chinese contact period. In general, vessels become shorter in height over time, leading to an oval-shaped body. Moreover, size significantly varies with shape for vessels before and after the European presence. Shorter vessels with a narrower opening tended to have a larger body diameter. This correlation between size and shape can be also observed during the Chinese period, but there is no relationship with height.

Our CV tests indicate that there are significant differences in shape standardization between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods. Pottery shape became more homogeneous and standardized after contact with Europeans and even more so after contact with the Chinese. Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. To determine if form changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf. @Kwak2015]. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909].

>>>>

Key questions in archaeological shape analysis normally involve measuring shape standardization over time, or between geographical areas. Standardization is often investigated using multivariate analysis of shape variables computed from landmark data, or coefficients of variation on associated metric data, especially for lithic assemblages. For example, @Archer2015's case study of stone points in Southern Africa suggests an increase in shape standardization over time that may relate to increased maintenance of finished points. @Buchanan2018 studied metric data and GMM to identify a more uniform base-shape of Folsom points than Clovis points across the western US [@Buchanan2019]. Also using metric data and GMM, @Smith2016 found standardized bases of fluted point in Alaska and northern Yukon that might indicate a risk management strategy to ensure the ease of replacement during long-distance travel. Other factors, such as low levels of cultural innovation in a small group, could also lead to an increase in standardization of point shapes [@Okumura2014]. To test the effectiveness of measuring standardization, @Birch2019 compared landmark-based GMM to traditional metric analysis with CV using European iron weapons as an example. They demonstrated that landmark-based GMM can capture more variation in not only overall shape, but also bilateral symmetry.

@@

As a standardized measure of the spread of data, coefficients of variation (CV) allows us to directly compare variation in samples measured with different units or means. This is useful to compare the degree of standardization for archaeological assembles and enables comparison of variation across different sample sizes [@Eerkens2001, pp.498]. Following @Eerkens2001 and @Roux2003, we take this as our measurement of standardization in vessel shape variables: lower CV values reflect higher standardization, and thus increased craft specialization in the community. Given that CV is robust for positive values due to the representation by a ratio scale, we normalized PC scores to a range between 1 and 10 before computation of CV.

@@

We also conducted size analysis by examining the body diameter of vessels to examine any differences in vessel size and their relationships with vessel shape. We measured body diameter directly from each physical vessel in the collection, and we focus on this metric because it is available for more vessels than any other metric. The body diameter of vessels from the Chinese period is larger than those from the two earlier periods, and the vessels before European contact have the smallest body diameter on average (Figure \@ref(fig:size-shape-regression-plots): A). To investigate standardization of vessel form reflected by both shape and size, we calculated CV for PC1 (shape) in relation to CV for body diameter (size). The result (Figure \@ref(fig:size-shape-regression-plots): B) presents a higher form standardization in the Chinese period according to smaller CV values compared to CV values from the other two phases. However, there is no big differences in form standardization between and after the European presence. To understand the relationship between shape and size, we created linear regression models for PCs and the measurement of body diameter, with an examination of Pearson's correlation coefficient and p-value (Figure \@ref(fig:size-shape-regression-plots): C). The results demonstrate that shape changes significantly with size in all phases indicated by moderate positive relationships (0.3 ≤ r ≤ 0.7) and small p-values (≤ 0.05), except for PC1 in the Chinese period and PC2 in the pre-European period. That is, the shorter vessels are larger in body diameter according to the positive correlation between PC1 and body diameter. However, vessels from the Chinese period do not show this pattern. On the contrary, the negative relationship between PC2 and body diameter suggests that vessels with a narrower neck and mouth tend to have a larger body diameter.

@@

```{r cv-metric-table-with-pc1-prep}
# combine with cv information 
size_pcs_cv_join <-
  jpgs_imported_coo_size_pcs_metric %>% 
  group_by(y) %>% 
  mutate(cv_body_diameter = round(raster::cv(Diameter1_Body), 2),
         cv_neck_diameter = round(raster::cv(Diameter1_Neck), 2),
         cv_rim_diameter = round(raster::cv(Diameter1_Rim), 2),
         Diameter1_Body_mean = round(mean(Diameter1_Body), 2)) %>% 
  mutate(phase = as.factor(y)) %>% 
  left_join(cv_pcs_all) %>% 
  left_join(pc1_df_with_filename) %>% 
  select(-y)
  #distinct(phase, cv_pc1, cv_pc2, cv_pc3, Diameter1_Body_mean, cv_body_diameter)
```

```{r boxplot-size-by-phase}
# prepare label
size_cv_df <-
  size_pcs_cv_join %>% 
  distinct(phase, Diameter1_Body_mean, cv_body_diameter, cv_pc1) %>% 
  mutate(cv_body_diameter_label = paste("CV =", cv_body_diameter, "%"),
         Diameter1_Body_mean_label = paste("mean=", Diameter1_Body_mean))

@@

```{r byplot-cvs-pc1-size}
library(ggrepel)
y_limits <- c(NA, 32)
byplot_cvs_pc1_diameter <-
  size_cv_df %>% 
  ggplot(aes(cv_body_diameter, cv_pc1)) +
  geom_point() +
  geom_text_repel(label = size_cv_df$phase, 
                  size = 2.5) + 
  labs(x = "CV of body diameter (%)",
       y = "CV of PC1\n(%)") +
  theme(legend.position = "none") +
  theme_bw() +
  theme(axis.text.x = element_text(size = rel(0.9)),
        axis.text.y = element_text(size = rel(0.9)))
```

@@

```{r plot-size-regression-together}
# plot boxplot and regression together 
left_column <-
  plot_grid(boxplot_size, byplot_cvs_pc1_diameter, 
            ncol = 1, rel_heights = c(1.1,1), 
            labels = c("A", "B"), label_size = 12 )

@@

ggsave(plot = size_boxplot_regression,
       here("analysis", "figures",
            "plot-size-boxplot-regression.jpg"),
       width = 180, 
       height = 100,
       dpi = 300,
       units = "mm")
```

```{r, size-shape-regression-plots, echo=FALSE, fig.cap= "A: Distribution of the body diameter of vessels with coefficients of variation by phases. B: Coefficients of variation of vessel shape represented by PC1 in relation to vessel size represetnted by body diameter, showing a more standardization in vessel form in the Chinese period. C: Correlation between shape variables (PC1 and PC2) and body diameter of vessles with Person's r and P-values by phases"}
knitr::include_graphics(here("analysis/figures/size-boxplot-regression.jpg"))
```

@@

Previous investigations at Kiwulan suggested an unequal distribution of prestige goods, trade ornaments in particular, following the appearance of Europeans [@Cheng2008; @Wang2011], which hinted at the emergence of social inequality within the Indigenous community. Ceramic vessel shape standardization was examined to measure craft specialization as a proxy for social differentiation [@Costin2001; @Junker1999]. The result of our MANOVA demonstrates a significant difference in shapes between the pre-European and Post-European periods, and between the pre-European and the Chinese periods. The average shape presents as a round body with a wide rim and neck before European contact, which shifts to a more oval-shaped body with narrower rim and neck after the European presence. Shape changes are more pronounced during the Chinese contact period. In general, vessels become shorter in height over time, leading to an oval-shaped body. This can be also supported by an increase in body diameter of vessels. Also, the correlation between shape and body diameter suggests that size significantly varies with vessel shape. Shorter vessels with a narrower opening tended to have a larger body diameter, indicative of a change in overall vessel form.

Our CV tests indicate that there are significant differences in shape standardization between the pre-European period and post-European period, and between the Chinese contact and either pre-European or post-European periods. Pottery shape became more homogeneous and standardized after contact with Europeans and even more so after contact with the Chinese. In addition, we found a larger and more standardized body diameter of vessels in the Chinese period. We notice the possibility of more human errors could be involved in hand-crafting when the size of artifacts increases [@Eerkens2001]. However, it is opposite for the ceramics in the Chinese period when we examining body diameter. This might support the intent to achieve a homogeneous form of vessels, but we still need to compare it with other metric variables to validate this claim. Compositional analysis shows that the clay pastes are similar, regardless of the increasing standardization of the pottery shape, reflecting continuity in the raw material sources. We can thus rule out changes in clay fabric as a factor in explaining changes in vessel shape. To determine if form changes might be related to changes in the function of pots at Kiwulan, we used geochemical methods to extract and identify lipids trapped in the fabric of potsherds to identify foods that may have contributed residues absorbed into the clay [cf. @Kwak2015]. Unfortunately, we did not obtain useful results due to extremely low lipid yields, which is probably due to the very thin, dense, and low porosity fabric of Kiwulan pottery. These physical characteristics of the clay offer limited spaces to trap and protect organic molecules from microbiological degradation [cf. @Evershed2008, pp. 909].
